Mechanical and relaxation-based detection of dipolar-interactions between spins in diamond

### 2020-11-17 Mes papiers/Méca CR/Meca_cross_SI.tex
okou

The plain lines are the locus of the planes orthogonal to the [100] directions (3 planes total). When the magnetic field is in these planes, every class is at resonance with another class, as can be seen in Fig. X.

@@

Finally the blue dots correspond to the path followed by the magnetic field in the experiment presented in Fig 4. of the main text, where we can see a degeneracy plane of the [110]$_\perp$ family being crossed.

@@

\subsection{$T_1$ Measurement}
\begin{figure}[!ht]
  \centering \scalebox{1.2}{\includegraphics{T1_soustraction_shema}}
  \caption{T$_1$ measurement protocol for a single class in an ensemble of NV centers. Green bars represent laser excitation, red bar represent microwave $\pi$ pulse for the selected class}
	\label{T1_protocol}
\end{figure}
The protocol to measure the spin lifetime of the NV centers, as presented in Fig 2. of the main text, is described in Fig. \ref{T1_protocol} . In the first sequence ($S_1$), the spins are initially polarized in their $m_s=\ket{0}$ state through a 1 ms green laser excitation pulse and then left to evolve in the dark for a variable dark time $\tau$. The spin state is finally read out with a 10 $\mu$s laser pulse, shorter than the polarization time of the spins.

The second sequence ($S_2$) uses the same parameters (polarization time, dark time and readout time) than the first sequence but adds a resonant microwave $\pi$ pulse on a transition of one of the four classes of NV$^-$ right before the readout pulse, exchanging the population of the $m_s=\ket{0}$ state with one of the $m_s=\ket{\pm 1}$ state.

By looking a the subtraction of both signal $S_1-S_2$, we can then measure the spin state population of a single class, and remove unwanted contributions to the photoluminescence, such as charge state transfer in the dark.

@@

@@

@@

We model the trap as a pure harmonical potential, both for the center of mass and for the rotational degrees of freedom of the diamond (REF qq part), with trapping frequencies $\omega_T \approx 1$ kHz that we can experimentally measure. If we focus on a single rotational degree, we can write the torque generated by the trap as $\Gamma_T^\theta=-K(\theta-\theta_{eq})$, where $K=I \omega_t^2$ is the stiffness of the trap, $I$ being the moment of inertia of the diamond.

@@

In our case, $\Gamma_{\mathrm{ext}}$ is the magnetic torque exerted by the NV$^-$ spins directly on the diamond, such that we can write $\Gamma_{\mathrm{ext}} = N_{NV} \langle \Gamma_{\mathrm{1 spin}} \rangle$  where $\langle \Gamma_{\mathrm{1 spin}} \rangle = \gamma_e \langle\hat{\mathbf S}\rangle \times \mathbf B \approx 10^{-28}$ Nm is the classical magnetic torque applied by one spin (see Section  \ref{Simu}).

By using the inertia moment formula of a sphere : $I=\frac{2}{5}m r^2$, we can then rewrite the angular displacement as $$ \delta\theta = \frac{\langle\Gamma_{\mathrm{1 spin}}\rangle [NV^-]}{\frac{2}{5}m_Cr^2\omega_T^2} \approx 10^{-3} rad$$ where $[NV^-] \approx 5 \cdot 10^{-6}$ (5 ppm) is the number of NV centers per atoms in the crystal, $m_C \approx 2 \cdot 10^{-26}$ kg is the average weight of a carbon atom (we assume that the bulk of the diamond weight comes from carbon atoms), $r= 7.5$ $\mu$m is the typical radius of our diamonds and $\omega_T = 1$ kHz is the typical value of the trap frequency.

@@

\subsection{Change of torque at the cross-relaxation}

@@

%Faire 2 lignes avec sur la première ligne l'ESR et le trajet sur la carte du champ mag
\begin{figure}[!ht]
  \centering \scalebox{0.32}{\includegraphics{CR_meca_22}}
  \caption{Mechanical detection of a degeneracy in a [100]$_\perp$ plane. a) Theoretical frequencies for the four $\ket{0} \to \ket{-1}$ transitions in dashed lines and mechanical ESR spectra vertically. b) Photoluminescence of the NV centers as a function of the scanning magnetic field (blue crosses) and gaussian fit (orange line). c) Simulated population in $\ket{m_s=0}$ for the stationary state with (plain) or without (dashed) taking into account the decrease of the T$_1$ induced by the cross-relaxations.
  d) Reflected signal off the diamond as a function of the scanning magnetic field (blue crosses) and double gaussian fit (orange line) e) Simulated torque applied by the spins on the diamond, with (plain) or without(dashed) taking into account the cross-relaxations.}
	\label{CR_22}
\end{figure}

Similarly to Fig. 3 of the main text, we managed to mechanically detect another type of NV-NV resonance. Fig. \ref{CR_22}a) shows the theoretical frequencies of the $\ket{0} \to \ket{-1}$ transitions for all four classes of NV and ESR spectra measured through the reflected laser light off the diamond for various magnetic field values. Unlike The experiment of the main text, this time all four classes of NV are at resonance with another class at B=23 G, telling us that we are crossing a [100]$_\perp$ plane instead of a [110]$_\perp$ one.

Fig. \ref{CR_22}b) shows the recorded photoluminescence of the NV centers during the magnetic field scan. As previously, we see a drop in the luminosity when the resonance occurs, that we fitted with a gaussian. The drop is slightly more pronounced in this case since all classes are depolarized, instead of only two in the previous experiment. This behavior is predicted by the simulation of the $\ket{m_s=0}$ population in the stationary state in Fig. \ref{CR_22}c).

Fig. \ref{CR_22}d) shows the signal of the reflected laser off the diamond, allowing us to measure angular displacement. This time we see a clear difference with the experiment of the main text, instead of a single drop centered on the resonance, we see two bumps on both sides of the resonance, fitted with two gaussians.
This is in accordance with the simulated torque in Fig. \ref{CR_22}e) where we can see a dispersive profile with an almost zero torque at the resonance. The reason we do not observe a change of sign in the experiment (with two positive bumps instead of a positive and a negative one) might be because of the strong non-linearity of our detection : if the signal initially correspond to a dark spot of the speckle, then a change in the motion of the diamond can only result in an increased signal. We also note a drop both in the predicted torque amplitude, and in the signal-to-noise ratio  of the angular measurement compared to the single degeneracy case, implying a weaker effect.

Our physical interpretation of the angular signal we observe, as well as the dispersive profile of the torque is that in this case, for symmetry reason, the magnetic torque generated by the four classes of NV is not modified when we are exactly at resonance, since all four classes are affected identically. What happens when being near, but not at, resonance is that the classes will not be identically depolarized : looking at Fig. \ref{CR_22}a), we can see that the two classes of higher frequency are always slightly closer to each other than the two classes of lower frequency (we can see the slope of the frequencies being a bit smaller for the two upper classes), resulting in a bit more depolarization for these two classes, except when they are exactly at resonance. This interpretation accounts both for the overall shape of the responses, and the comparatively weakness of the phenomenon.

>>>>

The plain lines are the locus of the planes orthogonal to the [100] directions (3 planes total). When the magnetic field is in these planes, every class is at resonance with another class, as can be seen in Fig. \ref{CR_22}.

@@

Finally the blue dots correspond to the path followed by the magnetic field in the experiment presented in Fig 3. of the main text, where we can see a degeneracy plane of the [110]$_\perp$ family being crossed.

@@

\subsection{$T_1$ Measurement}
\begin{figure}[!ht]
  \centering \scalebox{1.2}{\includegraphics{T1_soustraction_shema}}
  \caption{T$_1$ measurement protocol for a single class in an ensemble of NV centers. Green bars represent laser excitation, red bar represent microwave $\pi$ pulse for the selected class}
	\label{T1_protocol}
\end{figure}
The protocol to measure the spin lifetime of the NV centers, as presented in Fig 2. of the main text, is described in Fig. \ref{T1_protocol} . In the first sequence the spins are initially polarized in their $m_s=\ket{0}$ state through a 1 ms green laser excitation pulse and then left to evolve in the dark for a variable dark time $\tau$. The spin state is finally read out with a 10 $\mu$s laser pulse, shorter than the polarization time of the spins.

The second sequence uses the same parameters (polarization time, dark time and readout time) than the first sequence but adds a resonant microwave $\pi$ pulse on a transition of one of the four classes of NV$^-$ right before the readout pulse, exchanging the population of the $m_s=\ket{0}$ state with one of the $m_s=\ket{\pm 1}$ state.

Noting $S_1$ the signal given by the first sequence and $S_2$ the one given by the second sequence, then by looking at $S_1-S_2$ we can measure the spin state population of a single class, and remove unwanted contributions to the photoluminescence, such as charge state transfer in the dark (which would be common to the two signals and therefore subtracted).
In order to avoid low frequency noises or drift, we alternate both sequence when performing the measurement.

@@

\subsection{Signal drift for levitating particles}

Measurements on levitating diamonds have to be relatively short (few minutes at most) because of a slow drift on the particle orientation. The most likely origin of this drift is the loss of charges of the diamond due to photoionization by the laser, which changes the trapping conditions over time.

This drift is the reason why the photoluminescence scan on Fig.3 of the main text has a lower signal to noise ratio than the scan in Fig.1 of the main text, which was done on a static diamond for a few hours.

@@

\subsection{Origin of the magnetic torque and link to cross-relaxation}

\begin{figure}[ht]
\centering{ \scalebox{0.4} {\includegraphics{map_torque_1_classe}}}
  \caption{\textbf{a)} Skecth of an NV center aligned in the \textbf{z} direction in the presence of an eternal magnetic field \textbf{B} forming an angle $\theta$ with the NV axis, and resulting magnetization $\gamma_e \langle \mathbf{ \hat S} \rangle$ of the NV center
   \textbf{b)} Plotting of the longitudinal ($S_\parallel$ in dashed line) and transverse ($S_\perp$ in plain line) components of the spin operator average value, in $\hbar$ unit, as a function of $\theta$ and for a magnetic field amplitude $|\mathbf{B}|=100$ G.   
   \textbf{c)} Amplitude of the magnetic torque $\mathbf \Gamma = \gamma_e \langle\hat{\mathbf S}\rangle \times \mathbf B$ acting on a single spin as a function of $\theta$ for $|\mathbf{B}|=100$ G.
   \textbf{d)} Amplitude of the same magnetic torque in the crystalline basis with $\theta$ and $\phi$ being the polar and azimuthal angle with respect to the [100] direction.}
  		\label{Torque1classe}
\end{figure}

The magnetic torque responsible for the motion of the diamond fundamentally comes from the crystalline anisotropy of the NV centers and from the transverse field $B_\perp$ responsible for mixing the eigenstates in the stationary state. We will start by focusing on a single NV center.

As can be seen in Fig \ref{Torque1classe} b), when under green excitation and in the presence of an external magnetic field, the spins will acquire a magnetization $\gamma_e \langle\hat{\mathbf S}\rangle$ which, at the low magnetic fields ($<$ 200 G) we are working at, will be oriented at a 90$^\circ$ angle from the NV axis : $\langle \hat S_z \rangle \approx 0$ and $\langle \hat S_\perp \rangle \neq 0$ This magnetization vanish when the magnetic field is aligned with the NV center since there is no longer a transverse field responsible for the mixing of the eigenstates.

The magnetization of the NV center is therefore not aligned with the magnetic field, except when the field is also at a 90$^\circ$ angle from the NV axis, which means that the magnetic torque $\mathbf \Gamma = \gamma_e \langle\hat{\mathbf S}\rangle \times \mathbf B $ will be non-zero everywhere except when the field is aligned with the center, or in the plane normal to the direction of the center. We can describe each NV center as a paramgnetic defect with the anisotropic magnetic suceptibility $\chi = \begin{pmatrix}-\chi_\perp & 0 & 0\\ 0 & -\chi_\perp & 0 \\ 0 & 0 & 0  \end{pmatrix}$ in the $(\mathbf{x},\mathbf{y},\mathbf{z})$ basis where $\mathbf{z}$ is the orientation of the NV center.

The amplitude of the torque with respect to the magnetic field orientation (for an amplitude of 100 G) is represented in 1D in Fig \ref{Torque1classe} c) where we can see a behavior very close to $|sin(2\theta)|$, which is the formula (3) of the main text that we obtained through a perturbative approach. The same torque amplitude is represented in 2D in Fig \ref{Torque1classe} d) where the two dots correspond to the [111] direction, when the magnetic field is aligned with the centers, and the curvy line is the plane $[111]_\perp$. Important to note is the maximum torque value $1\cdot 10^{-26}$ Nm for a single spin.

\begin{figure}[ht]
\centering{ \scalebox{0.4} {\includegraphics{map_torque_4_classe}}}
  \caption{Magnetic torque amplitude for 4 spins, one in each possible orientation, as a function of $\theta$ and $\phi$, the polar and azimuthal angle with respect to the [100] direction. \textbf{a)} Not taking into account the CR between NV centers. \textbf{b)} Taking into account the CR between NV centers.}
  		\label{Torque4classes}
\end{figure}

Fig \ref{Torque4classes} a) represents the same map but for four NV centers, one in each possible orientation. We can see that the maximum torque actually decreased to $3\cdot 10^{-28}$ Nm even though we increased the number of NV centers by four. This is due to the directional averaging of the torque generated by the four centers. The torque per NV center is decreased by more than two order of magnitude when taking the directional averaging in consideration.

Finally, Fig \ref{Torque4classes} b) shows the same map, this time taking into account the modification of the spins lifetime due to cross-relaxations. There are two things to note here : 
\begin{enumerate}
\item The maximum torque has increased by an order of magnitude compared to the previous case, up to $3\cdot 10^{-27}$ Nm for four spins, or about $10^{-27}$ Nm per spin. Qualitatively, this is because cross-relaxation will lower the torque contribution of specific classes (the ones that get depolarized), meaning the the end result is closer to the single spin case (there is less directional averaging)
\item The change in magnetic torque is resonant, it happens only when different classes are brought to resonance (we can see in Fig \ref{Torque4classes} b) the [110]$_\perp$ planes that were drawn in Fig. \ref{cristallo}), meaning that the change in signal when scanning a magnetic field will be much sharper.
\end{enumerate}

@@

We model the trap as a pure harmonical potential, both for the center of mass and for the rotational degrees of freedom of the diamond (REF ?), with trapping frequencies $\omega_T \approx (2\pi) \cdot 1$ kHz that we can experimentally measure. If we focus on a single rotational degree, we can write the torque generated by the trap as $\Gamma_T=-K(\theta-\theta_{eq})$, where $K=I \omega_t^2$ is the stiffness of the trap, $I$ being the moment of inertia of the diamond.

@@

In our case, $\Gamma_{\mathrm{ext}}$ is the magnetic torque exerted by the NV$^-$ spins on the diamond, such that we can write $\Gamma_{\mathrm{ext}} = N_{NV} \langle \Gamma_{\mathrm{1 spin}} \rangle$  where $\langle \Gamma_{\mathrm{1 spin}} \rangle = \gamma_e \langle\hat{\mathbf S}\rangle \times \mathbf B \approx 10^{-27}$ Nm is the expected magnetic torque applied by one spin (see Fig \ref{Torque4classes}).

By using the inertia moment formula of a sphere : $I=\frac{2}{5}m r^2$, we can then rewrite the angular displacement as $$ \delta\theta = \frac{\langle\Gamma_{\mathrm{1 spin}}\rangle n(NV^-)}{\frac{2}{5}m_Cr^2\omega_T^2} \approx 10^{-4} rad$$ where $n(NV^-) \approx 5 \cdot 10^{-6}$ (5 ppm) is the number of NV centers per atoms in the crystal, $m_C \approx 2 \cdot 10^{-26}$ kg is the average weight of a carbon atom (we assume that the bulk of the diamond weight comes from carbon atoms), $r= 7.5$ $\mu$m is the typical radius of our diamonds and $\omega_T = 6.3\cdot 10^3$ rad/s is the typical value of the trap angular frequency.

@@

@@

%Faire 2 lignes avec sur la première ligne l'ESR et le trajet sur la carte du champ mag
\begin{figure}[!ht]
  \centering \scalebox{0.4}{\includegraphics{22-total}}
  \caption{Mechanical detection of a degeneracy in a [100]$_\perp$ plane.\textbf{a)} Path of the magnetic field along the scan (red dots) in the ($\theta , \phi$) basis with respect to the [100] direction. [100]$_\perp$ family of planes in plain blue lines, [110]$_\perp$ family of planes in orange dashed lines.\textbf{b)} Theoretical frequencies for the four $\ket{0} \to \ket{-1}$ transitions in dashed lines and experimental ESR spectra vertically. \textbf{c)} Photoluminescence of the NV centers as a function of the scanning magnetic field (blue crosses) and gaussian fit (orange line). \textbf{d)} Simulated population in $\ket{m_s=0}$ for the stationary state with (plain) or without (dashed) taking into account the decrease of the T$_1$ induced by the cross-relaxations.
  \textbf{e)} Reflected signal off the diamond as a function of the scanning magnetic field (blue crosses) and double gaussian fit (orange line) \textbf{f)} Simulated torque applied by the spins on the diamond, with (plain) or without(dashed) taking into account the cross-relaxations.}

	\label{CR_22}
\end{figure}

Similarly to Fig. 3 of the main text, we managed to mechanically detect another type of NV-NV resonance. Fig. \ref{CR_22}b) shows the theoretical frequencies of the $\ket{0} \to \ket{-1}$ transitions for all four classes of NV and ESR spectra measured through the reflected laser light off the diamond for various magnetic field values. Unlike The experiment of the main text, this time all four classes of NV are at resonance with another class at B=23 G, telling us that we are crossing a [100]$_\perp$ plane instead of a [110]$_\perp$ one, as can be seen on fig \ref{CR_22}a)

Fig. \ref{CR_22}c) shows the recorded photoluminescence of the NV centers during the magnetic field scan. As previously, we see a drop in the luminosity when the resonance occurs, that we fitted with a gaussian. The drop is slightly more pronounced in this case since all classes are depolarized, instead of only two in the previous experiment. This behavior is predicted by the simulation of the $\ket{m_s=0}$ population in the stationary state in Fig. \ref{CR_22}d).

Fig. \ref{CR_22}e) shows the signal of the reflected laser off the diamond, allowing us to measure angular displacement. This time we see a clear difference with the experiment of the main text, instead of a single drop centered on the resonance, we see two bumps on both sides of the resonance, fitted with two gaussians.
This is in accordance with the simulated torque in Fig. \ref{CR_22}f) where we can see a dispersive profile with an almost zero torque at the resonance. The reason we do not observe a change of sign in the experiment (with two positive bumps instead of a positive and a negative one) might be because of the strong non-linearity of our detection : if the signal initially correspond to a dark spot of the speckle, then a change in the motion of the diamond can only result in an increased signal. We also note a drop both in the predicted torque amplitude, and in the signal-to-noise ratio  of the angular measurement compared to the single degeneracy case, implying a weaker effect.

Our physical interpretation of the angular signal we observe, as well as the dispersive profile of the torque is that in this case, for symmetry reason, the magnetic torque generated by the four classes of NV is not modified when we are exactly at resonance, since all four classes are affected identically. What happens when being near, but not at, resonance is that the classes will not be identically depolarized : looking at Fig. \ref{CR_22}b), we can see that the two classes of higher frequency are always slightly closer to each other than the two classes of lower frequency (we can see the slope of the frequencies being a bit smaller for the two upper classes), resulting in a bit more depolarization for these two classes, except when they are exactly at resonance. This interpretation accounts both for the overall shape of the responses, and the comparatively weakness of the phenomenon.

### 2020-11-17 abstract/IQFA 2020/Abstract_Pellet-Mary.tex
demenagement
okou

\begin{abstract}
\normalsize
The negatively charged nitrogen vacancy center (NV$^-$) is a well established quantum system, employed both for applications and fundamental research, thanks to its good coherence properties, long lifetimes and most importantly their ability to be optically polarized.
We have recently found new ways to use these properties of NV centers in order to study the dipolar interactions between ensemble of spins in diamond, including NV centers as well as other spin defects.

On the one hand, we have observed cross-relaxations (CR) between NV centers and other defects (see Fig. 1), namely VH$^-$ \cite{VH}, War1 (first defect found by EPR at the university of Warwick, still chemically unknown) and $^{13}$C First-shell modified NV. None of these defects had been observed through CR before. This is a first step toward hyper-polarization of new spin defects in diamond.

On the other hand, we have used the mechanical detection of a levitating diamond \cite{nature} to measure the torque applied by the NV centers on the diamond when two classes of NV with different orientation are brought into resonance(see Fig. 2). This effect is due to the dipolar-mediated modification of the T$_1$ of NV centers happening with dense ensembles.\cite{Lukin}. This paves the way toward the mechanical detection of other spin impurities in diamond, as well as the observation of the Einstein-de Haas effect with paramagnetic systems. \cite{Meriles}

@@

Figure 1 : Photoluminescence change while scanning the magnetic field along the crystalline [100] direction. Three dips are observed for the CR with $^{13}$C-NV, VH$^-$ and War1.

@@

Figure 2 : Angular position of the diamond while scanning B$_{\mathrm{Scan}}$ without a microwave. The rotation of the diamond at 17 G corresponds to the resonance between two classes of NV
\end{abstract}

>>>>

\begin{abstract}
\normalsize
The electronic spin of the negatively charged nitrogen vacancy center (NV$^-$) is employed both for applications and fundamental research, thanks to its good coherence properties, long lifetimes and most importantly their ability to be optically polarized.
We have recently found new ways to use these properties of NV centers in order to study the dipolar interactions between ensemble of spins in diamond, including NV centers as well as other spin defects.

On the one hand, we have observed cross-relaxations (CR) between NV centers and other defects (see Fig. 1), namely VH$^-$ \cite{VH}, War1 (first defect found by EPR at the university of Warwick, still chemically unknown) and $^{13}$C-NV pair. None of these defects had been observed through CR before. This observation is a first step toward hyper-polarization of new spin defects in diamond.

On the other hand, we have used a levitating diamond \cite{nature} to measure the torque applied by the NV centers on the diamond when two classes of NV with different orientation are brought into resonance (see Fig. 2) because of the dipolar-mediated modification of the T$_1$ of NV in dense ensembles\cite{Lukin}. This demonstration paves the way toward the mechanical detection of other spin impurities in diamond, as well as the observation of the Einstein-de Haas effect with paramagnetic systems\cite{Meriles}.

@@

Figure 1 : \textit{Relaxation-based detection of dipolar interaction.} \textbf{Left} : Sketch of the various spins interacting. \textbf{Right} : Photoluminescence change while scanning the magnetic field along the crystalline [100] direction. Three dips are observed for the CR with $^{13}$C-NV, VH$^-$ and War1.

@@

Figure 2 : \textit{Mechanical detection of dipolar interactions.} \textbf{Left} : Sketch of the experimental setup. \textbf{Right :} Angular position of the diamond while scanning B$_{\mathrm{Scan}}$ without a microwave. The rotation of the diamond at 17 G corresponds to the resonance between two classes of NV.
\end{abstract}
